---
title: "Mobility Before and After COVID-19"
author: "Juthi Dewan, Son Phan"
output:
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(sf)
library(lubridate)
library(survminer)
library(survival)

theme_map <- theme_void()

taxi_zones <- st_read("../shp/taxi_zones.shp")
load(file = "../data/rides_20_h1.RData")
load("../data/rides_19_q1.RData")
load("../data/rides_19_q2.RData")
load("../data/rides_19_q3.RData")
load("../data/rides_19_q4.RData")


rides_20_h1clean <- rides_20_h1 %>%
	mutate(year = year(tpep_dropoff_datetime),
				 month = month(tpep_dropoff_datetime),
				 week = week(tpep_dropoff_datetime),
				 day = yday(tpep_dropoff_datetime), 
				 weekday = weekdays(tpep_dropoff_datetime),
				 weekend = weekday %in% c("Saturday", "Sunday"),
				 trip_time = as.numeric(tpep_dropoff_datetime - tpep_pickup_datetime),
				 speed = trip_distance/trip_time) %>%
	filter(month <= 6, trip_time > 0)

rides_19_q1_sample <- rides_19_q1 %>%
	sample_n(size = 7000000) %>%
	mutate(year = year(tpep_dropoff_datetime),
				 month = month(tpep_dropoff_datetime),
				 week = week(tpep_dropoff_datetime),
				 day = yday(tpep_dropoff_datetime), 
				 weekday = weekdays(tpep_dropoff_datetime),
				 trip_time = as.numeric(tpep_dropoff_datetime - tpep_pickup_datetime),
				 weekend = weekday %in% c("Saturday", "Sunday"),
				 speed = trip_distance/trip_time) %>%
	filter(month <= 6, trip_time > 0)

rides_19_q2_sample <- rides_19_q2 %>%
	sample_n(size = 7000000) %>%
	mutate(year = year(tpep_dropoff_datetime),
				 month = month(tpep_dropoff_datetime),
				 week = week(tpep_dropoff_datetime),
				 day = yday(tpep_dropoff_datetime), 
				 weekday = weekdays(tpep_dropoff_datetime),
				 trip_time = as.numeric(tpep_dropoff_datetime - tpep_pickup_datetime),
				 weekend = weekday %in% c("Saturday", "Sunday"),
				 speed = trip_distance/trip_time) %>%
	filter(month <= 6, trip_time > 0)

rides_19_q3_sample <- rides_19_q3 %>%
	sample_n(size = 7000000) %>%
	mutate(year = year(tpep_dropoff_datetime),
				 month = month(tpep_dropoff_datetime),
				 week = week(tpep_dropoff_datetime),
				 day = yday(tpep_dropoff_datetime), 
				 weekday = weekdays(tpep_dropoff_datetime),
				 trip_time = as.numeric(tpep_dropoff_datetime - tpep_pickup_datetime),
				 weekend = weekday %in% c("Saturday", "Sunday"),
				 speed = trip_distance/trip_time) %>%
	filter(month <= 12, trip_time > 0)

rides_19_q4_sample <- rides_19_q4 %>%
	sample_n(size = 7000000) %>%
	mutate(year = year(tpep_dropoff_datetime),
				 month = month(tpep_dropoff_datetime),
				 week = week(tpep_dropoff_datetime),
				 day = yday(tpep_dropoff_datetime), 
				 weekday = weekdays(tpep_dropoff_datetime),
				 trip_time = as.numeric(tpep_dropoff_datetime - tpep_pickup_datetime),
				 weekend = weekday %in% c("Saturday", "Sunday"),
				 speed = trip_distance/trip_time) %>%
	filter(month <= 12, trip_time > 0)

rides_19_h1_sample <- bind_rows(rides_19_q1_sample, rides_19_q2_sample)
rides_19_h2_sample <- bind_rows(rides_19_q3_sample, rides_19_q4_sample)
```

## Distribution of rides over time
```{r}
ride_ct_weeks_20 <- rides_20_h1clean %>%
	count(day) %>%
	mutate(total_rides = nrow(rides_20_h1clean),
				 ride_proportion = n/total_rides)

ride_ct_weeks_19 <- rides_19_h1_sample %>%
	count(day) %>%
	mutate(total_rides = nrow(rides_19_h1_sample),
				 ride_proportion = n/total_rides)

ggplot() +
	geom_line(data = ride_ct_weeks_20, mapping = aes(x = day, y = ride_proportion, color="2020")) + 
	geom_line(data = ride_ct_weeks_19, mapping = aes(x = day, y = ride_proportion, color="2019"))
```

## Distribution of rides over space
```{r, warning=FALSE}
# Seems to be outliers in Manhattan obviously
rides_20_h1clean %>%
	filter(DOLocationID == 230) %>%
	count(PULocationID) %>%
	mutate(total_rides = sum(n),
				 ride_proportion = n/total_rides,
				 log_rides = log(n)) %>%
	left_join(x = ., y = taxi_zones, by = c("PULocationID" = "LocationID")) %>%
	ggplot(mapping = aes(geometry = geometry)) +
	geom_sf(mapping = aes(fill = log_rides)) + 
	theme_map +
	scale_fill_viridis_c() +
	theme(legend.position="bottom") + 
	guides(fill = guide_colourbar(barwidth = 20)) + ggtitle("2020 H1")
```

## Comparing changes in trip_time between 2019 H1 and 2020 H1
Only consider the top 50-ish origins amongst the 2 years. 
```{r}
top_origins_20 = rides_20_h1clean %>%
	filter(DOLocationID == 230) %>%
	count(PULocationID) %>% 
  arrange(-n) %>%
  filter(row_number() <= 50) %>% pull(PULocationID)

top_origins_19 = rides_19_h1_sample %>%
	filter(DOLocationID == 230) %>%
	count(PULocationID) %>% 
  arrange(-n) %>%
  filter(row_number() <= 50) %>% pull(PULocationID)

top_origins = union(top_origins_20, top_origins_19)

rides_20_h1_ts = rides_20_h1clean %>%
	filter(DOLocationID == 230,
	       PULocationID %in% top_origins) %>%
  group_by(PULocationID) %>%
  summarise(
    n_trips = n(),
    mean_trip_time = mean(trip_time),
    median_trip_time = median(trip_time),
    log_mean_trip_time = log(mean_trip_time),
    log_median_trip_time = log(median_trip_time),
    mean_speed = mean(speed),
    median_speed = median(speed)
  )

rides_19_h1_ts = rides_19_h1_sample %>%
	filter(DOLocationID == 230,
	       PULocationID %in% top_origins) %>%
  group_by(PULocationID) %>%
  summarise(
    mean_trip_time = mean(trip_time),
    median_trip_time = median(trip_time),
    log_mean_trip_time = log(mean_trip_time),
    log_median_tripl_time = log(median_trip_time),
    mean_speed = mean(speed),
    median_speed = median(speed)
  ) 

rides_h1_ts_delta = left_join(x = rides_19_h1_ts, y = rides_20_h1_ts, by = c("PULocationID" = "PULocationID")) %>%
  mutate(mean_trip_time_delta = mean_trip_time.y - mean_trip_time.x)

rides_h1_ts_delta %>%
  filter(!(abs(mean_trip_time_delta - mean(mean_trip_time_delta)) > 2*sd(mean_trip_time_delta))) %>%
	left_join(x = ., y = taxi_zones, by = c("PULocationID" = "LocationID")) %>%
	ggplot(mapping = aes(geometry = geometry)) +
	geom_sf(mapping = aes(fill = mean_trip_time_delta)) + 
	theme_map +
	scale_fill_viridis_c() +
	theme(legend.position="bottom") + 
	guides(fill = guide_colourbar(barwidth = 20)) + ggtitle("2020 Minus 2019 Trip Times")
```

## Preliminary Model notes
* model ride times
* control distance
* airport label
* weekday seasonality
* 1, 132, 138 are airports (control?)
* variation around outer ring
* are ride times dictated by lockdown requirements
* focus on a single route
* year:month interaction
* kaplan meier where strata by pickup zone
* in general from same origin to same destination, do trips get faster from A to B across 2020? What about compared to 2019?
<<<<<<< HEAD
* 2019/2020 from origin 230, single route, survfit; log-rank between 2019vs.2020
* Surv(trip_time) ~ factor(PULocationID) + factor(PULocationID):year + year
* by month in 2020
* The farther pickup locations might have a bigger negative trip_time change. This is intuitive, how can we test this?

## Simple Model - Just by Year
```{r}
rides_20_h1_top_origins <- rides_20_h1clean %>%
  filter(DOLocationID == 230, PULocationID %in% top_origins)

rides_19_h1_top_origins <- rides_19_h1_sample %>%
  filter(DOLocationID == 230, PULocationID %in% top_origins)

rides_h1_top_origins = bind_rows(rides_20_h1_top_origins, rides_19_h1_top_origins) %>%
  filter(year %in% c(2019, 2020)) %>%
  mutate(PULocationID = factor(PULocationID),
         year = factor(year))

## Weibull
trip_time_year_weibull = survreg(formula = Surv(trip_time) ~ factor(year), dist = "weibull", data = rides_h1_top_origins)
trip_time_year_weibull %>% summary()

wei_19_trip_time_SC = function(x) pweibull(q = x, shape = 1/trip_time_year_weibull$scale, scale = exp(trip_time_year_weibull$coefficients[1] + trip_time_year_weibull$coefficients[2]*0), lower.tail = FALSE)

wei_20_trip_time_SC = function(x) pweibull(q = x, shape = 1/trip_time_year_weibull$scale, scale = exp(trip_time_year_weibull$coefficients[1] + trip_time_year_weibull$coefficients[2]*1), lower.tail = FALSE)

## Log-normal
trip_time_year_lnorm = survreg(formula = Surv(trip_time) ~ factor(year), dist = "lognormal", data = rides_h1_top_origins)
trip_time_year_lnorm %>% summary()

lnorm_19_trip_time_SC = function(x) plnorm(q = x, meanlog = trip_time_year_lnorm$coefficients[1] + trip_time_year_lnorm$coefficients[2]*0, sdlog = trip_time_year_lnorm$scale, lower.tail = FALSE)

lnorm_20_trip_time_SC = function(x) plnorm(q = x, meanlog = trip_time_year_lnorm$coefficients[1] + trip_time_year_lnorm$coefficients[2]*1, sdlog = trip_time_year_lnorm$scale, lower.tail = FALSE)

## Cox PH
trip_time_year_coxPH = coxph(formula = Surv(trip_time) ~ factor(year), data = rides_h1_top_origins)
trip_time_year_coxPH %>% summary()

plot(survfit(trip_time_year_coxPH, newdata = data.frame(year = c("2019" , "2020"))), xlab = "Seconds", ylab = "Survival Proportion", col=1:2, xlim=c(0, 3000))
curve(lnorm_19_trip_time_SC, lty=2, add=TRUE)
curve(lnorm_20_trip_time_SC, col=2, lty=2, add=TRUE)

plot(survfit(trip_time_year_coxPH, newdata = data.frame(year = c("2019" , "2020"))), xlab = "Seconds", ylab = "Survival Proportion", col=1:2, xlim=c(0, 3000))
curve(wei_19_trip_time_SC, lty=2, add=TRUE)
curve(wei_20_trip_time_SC, col=2, lty=2, add=TRUE)
```

## Complex Model - Year AND Zone
Question: If trips got FASTER in 2020 relative to 2019, did some zones speed up more? did any zones get actually slower? Can we discern a pattern to these changes in space?

Takes a long time to run; be careful before running. There's lots of output that we're gonna have to parse through and kind of summarize in a way. One thing I thought about was maybe testing another variable like distance and seeing if some of the covariates become insignificant. This would tell us if there is another explanatory variable confounding with simply sorting out PULocationID's.

```{r}
## Weibull
trip_time_yearzone_weibull = survreg(formula = Surv(trip_time) ~ year*PULocationID, dist = "weibull", data = rides_h1_top_origins)
trip_time_yearzone_weibull %>% summary()
```

```{r}
#2020 trip duration in mins
Duration_2020 <-
  rides_20_h1clean%>%
  mutate(Pickup = tpep_pickup_datetime,Dropoff = tpep_dropoff_datetime) %>%
  select(-c(tpep_pickup_datetime, tpep_dropoff_datetime)) %>%
  mutate(Duration = difftime(Dropoff, Pickup, units = "mins"))

#2019 trip duration in mins
Duration_2019h1 <-
  rides_19_h1_sample%>%
  mutate(Pickup = tpep_pickup_datetime,Dropoff = tpep_dropoff_datetime) %>%
  select(-c(tpep_pickup_datetime, tpep_dropoff_datetime)) %>%
  mutate(Duration = difftime(Dropoff, Pickup, units = "mins"))

Duration_2019h2 <-
  rides_19_h2_sample%>%
  mutate(Pickup = tpep_pickup_datetime,Dropoff = tpep_dropoff_datetime) %>%
  select(-c(tpep_pickup_datetime, tpep_dropoff_datetime)) %>%
  mutate(Duration = difftime(Dropoff, Pickup, units = "mins"))
```

```{r}
sample_2020 <- Duration_2020 %>%
	sample_n(size = 10000) 

sample_2019h1 <- Duration_2019h1 %>%
	sample_n(size = 10000) 

sample_2019h2 <- Duration_2019h2 %>%
	sample_n(size = 10000) 
```

```{r}
#weekly trip duration mean
weeklyAvg20 <-
  sample_2020 %>%
  group_by(week) %>%
  mutate(weeklyAvg = mean(Duration))

weeklyAvg19h1 <-
  sample_2019h1 %>%
  group_by(week) %>%
  mutate(weeklyAvg = mean(Duration))
```

```{r}
#trip duration by week in 2020 vs 2019
ggplot() +
	geom_line(data = weeklyAvg20, mapping = aes(x = week, y = weeklyAvg, color="2020")) + 
	geom_line(data = weeklyAvg19h1, mapping = aes(x = week, y = weeklyAvg, color="2019")) + ylab("Weekly Avg. Trip Time (Min)")
```

```{r}
KM2020 = survfit( Surv(Duration) ~ month , data=sample_2020, )
KM2019 = survfit( Surv(Duration) ~ month , data=sample_2019h1)

plot(KM2019, xlim=c(0,6), ylim=c(0.55,1), col=1:6)
plot(KM2020, xlim=c(0,6), ylim=c(0.55,1), col=1:6)
```

```{r}
#lm to see the relationship between distance and duration in 2019 vs 2020
lm( Duration ~ trip_distance, data=sample_2020)
lm( Duration ~ trip_distance, data=sample_2019h1)
```

# ```{r}
# #IGNORE THIS
# #trip duration mean by distance
# dur_dist_20 <-
#   sample_2020 %>%
#   group_by(trip_distance) %>%
#   mutate(mean_dur = mean(Duration))
# 
# dur_dist_19h1 <-
#   sample_2019h1 %>%
#   group_by(trip_distance) %>%
#   mutate(mean_dur = mean(Duration))
# 
# # trip distance and duration
# ggplot() +
# 	geom_line(data = dur_dist_20, mapping = aes(x = trip_distance, y = mean_dur, color="2020")) + 
# 	geom_line(data = dur_dist_19h1, mapping = aes(x = trip_distance, y = mean_dur, color="2019"))
# ```
